### 2020-08-09 Manuscript.Rmd
light edits

Gene expression is an important component of phenotypic diversity within and among units of biological organization, including cells, tissues, organs, and species [@Brawand:2011du; @gilad2005multi; @king1975evolution; @rifkin2003evolution]. There are a number of ways in which gene expression can change, including in the relative magnitude of expression, or temporal or spatial shifts in expression [@carroll2005evolution; @romero2012comparative; @wray2003evolution]. With high throughput RNA-seq data, we are able to quantify the relative expression level of a large number of genes within and among species, homologous organs, and developmental timepoints. Based primarily on qualitative distance measures and principal component analysis (PCA) of expression across 1:1 orthologs, most studies suggest that gene expression is more conserved among homologous organs of the same type across different species than among different organs within the same species; suggesting that molecular, cellular and developmental pathways are highly conserved among species [@Brawand:2011du; @breschi2016gene; @cardoso2019gene; @clarke2017evolutionary; @gilad2015reanalysis; @khaitovich2004neutral; @merkin2012evolutionary; @10.1093/molbev/msz212]. Other studies found the opposite result [@lin2014comparison; @pankey2014predictable; @tschopp2014relative; @yang2013organ], however re-analyses contest these findings [@gilad2015reanalysis; @gu2015understanding; @sudmant2015meta]. These global patterns of gene expression, however, fail to account for gene-specific patterns, which may vary in organ and species specificity, and clustering patterns may be driven by the behavior of a small subset of genes [@breschi2016gene].  In this study, we take a gene-centric approach using phylogenetic comparative methods (PCMs) to compare gene expression patterns between different siphonophore zooids, and one specialized tissue, and map expression values directly onto gene trees, in order to investigate evolutionary changes gene expression of different tissues across branches in gene trees.

@@

The next component of variation we considered was biological variation among sampled colonies. Within species, the greatest variance was among zooids/tissues, as opposed to between biological replicates of the same zooid/tissue (Figs. \@ref(fig:pca-diphyes)--\@ref(fig:pca-physalia)), with some exceptions, such as a developing palpon replicate in *Nanomia bijuga* (Fig. \@ref(fig:pca-nanomia)), and a developing gastrozooid and male gonodendra in *Frillagalma vityazi* (Fig. \@ref(fig:pca-Frillagalma)). Specimens were collected in the wild at different depths and over different time periods, but despite these potentially confounding environmental factors, the major variation we observe is among zooids/specialized tissues within species.

@@

@@

Within the four best sampled species , we also identified `r x$short_blast_hit %>% unique() %>% length()` differentially expressed putative transcription factors (`r length(Bargmannia_TF)` out of `r y[[2]] %>% sum()` in *Bargmannia elongata*, `r length(Frillagalma_TF)` out of `r y[[4]] %>% sum` identified in *Frillagalma vityazi*, `r length(Agalma_TF)` out of `r y[[1]] %>% sum()` in *Agalma elegans* and `r length(Nanomia_TF)` out of `r y[[5]] %>% sum()` identified  in *Nanomia bijuga*). Many of these transcription factors are significantly differentially expressed in several zooids regardless of developmental stage (both mature and developing zooids), and a subset were found to be significantly differentially expressed only in particular zooids and developmental stages (\@ref(fig:transcription-factors)).

@@

summary %>% dplyr::mutate(
    Species= cell_spec(Species, italic= TRUE)
) %>% kable(caption="Number of genes per  species (library size), number of sampled zooids (at least two  replicates, including developing and mature stages), total number of genes in gene trees (there can be multiple genes in a gene tree), and number of unique gene trees containing genes from this species.", caption.short ="Summary statistics for each of the sampled species, with details on the total number of genes in gene trees, and number of unique gene trees containing genes from this species.",booktabs=TRUE, align = c("l","c","c","c","c"), escape = FALSE) %>% kable_styling(full_width = F, font_size = 11) %>%
  kableExtra::kable_styling(latex_options = "hold_position") %>%
  column_spec(2:5, width = "5em")

@@

return(p)
}
```
In order to compare expression patterns for each zooid across siphonophore species, we used transcriptome and genome data from `r species_tree$tip.label %>% length()` cnidarian species to generate a total of `r gene_trees %>% length()` gene trees, of which `r gene_trees_ace_calibrated[!is.na(gene_trees_ace_calibrated)] %>% length()` gene trees passed filtering criteria. We used Orthofinder [@emms2019orthofinder] to reconcile the species tree and the gene trees and identify nodes that represent speciation and duplication events (Fig. \@ref(fig:methods-comp), step 1). Pruned and unpruned calibrated gene tree files are available as supplementary data. The number of genes represented in these gene trees is shown in table \@ref(tab:summarize-libraries). The internal nodes on these gene trees consist of `r nodes_ace %>% filter(is.na(species) & Ev=="S") %>% nrow()` speciation events and `r nodes_ace %>% filter(is.na(species) & Ev=="D") %>% nrow()` duplication events. Expression values, as transcripts per million 10K (TPM10K, see methods), were mapped to the tips of the gene trees (Fig. \@ref(fig:methods-comp), step 2), and maximum likelihood trait values were reconstructed at the nodes (Fig. \@ref(fig:methods-comp), step 3). Changes in expression across branches were calculated based on the difference between parent and child nodes, scaled by the branch length (Fig. \@ref(fig:methods-comp), step 4) [@dunn2018pairwise]. We focused on expression in a subset of zooids and tissues that are common across the sampled species: gastrozooids (developing and mature), nectophores (developing), palpons (mature), male and female gonodendra, and the pneumatophore. An exemplar gene tree, for Wnt family genes, with expression values from two zooids mapped at the nodes and tips is shown in figure \@ref(fig:wnt-plot).

@@

For comparitive expression analyses, expression values were normalized using a method we are calling transcripts per million 10K (TPM10K). For gene $i$ of a given species, TPM is typically calculated as [@li2009rna]:

@@

This work was supported by the National Science Foundation (DEB-1256695 and the Waterman Award). CM was also supported in part by a RI-EPSCoR Fellowship, NSF EPS-1004057. Analyses were conducted with computational resources and services at the Center for Computation and Visualization at Brown University, supported in part by the NSF EPSCoR EPS-1004057 and the State of Rhode Island. Analyses were also conducted with computational resources and services XXXXXXX Yale CCV

>>>>

Gene expression is an important component of phenotypic diversity within and among units of biological organization, including cells, tissues, organs, and species [@Brawand:2011du; @gilad2005multi; @king1975evolution; @rifkin2003evolution]. There are a number of ways in which gene expression can change, including in the relative magnitude of expression, or temporal or spatial shifts in expression [@carroll2005evolution; @romero2012comparative; @wray2003evolution]. With high throughput RNA-seq data, it is possible to quantify the relative expression level of a large number of genes within and among species, homologous organs, and developmental timepoints. Based primarily on qualitative distance measures and principal component analysis (PCA) of expression across 1:1 orthologs, most studies suggest that gene expression is more conserved among homologous tissues of the same type across different species than among different tissues within the same species; suggesting that molecular, cellular and developmental pathways are highly conserved among species [@Brawand:2011du; @breschi2016gene; @cardoso2019gene; @clarke2017evolutionary; @gilad2015reanalysis; @khaitovich2004neutral; @merkin2012evolutionary; @10.1093/molbev/msz212]. Other studies found the opposite result [@lin2014comparison; @pankey2014predictable; @tschopp2014relative; @yang2013organ], however re-analyses contest these findings [@gilad2015reanalysis; @gu2015understanding; @sudmant2015meta]. These global patterns of gene expression, however, fail to account for gene-specific patterns, which may vary in tissue and species specificity, and clustering patterns may be driven by the behavior of a small subset of genes [@breschi2016gene].  In this study, we take a gene-centric approach using phylogenetic comparative methods (PCMs) to compare gene expression patterns between different siphonophore zooids, and one specialized tissue, and map expression values directly onto gene trees, in order to investigate evolutionary changes gene expression of different tissues across branches in gene trees.

@@

The second component of variation we considered was biological variation among sampled colonies. Specimens were collected in the wild at different depths and over different time periods, but despite these environmental factors there was remarkably little variation among sampled colonies. Some samples did show greater variation across replciates, such as a developing palpon replicate in *Nanomia bijuga* (Fig. \@ref(fig:pca-nanomia)), and a developing gastrozooid and male gonodendra in *Frillagalma vityazi* (Fig. \@ref(fig:pca-Frillagalma)).

@@

The third component of variation we considered was among zooids/specialized tissues within species. This was the greatest component of variation (Figs. \@ref(fig:pca-diphyes)--\@ref(fig:pca-physalia)).

@@

Within the four best sampled species, we also identified `r x$short_blast_hit %>% unique() %>% length()` differentially expressed putative transcription factors (`r length(Bargmannia_TF)` out of `r y[[2]] %>% sum()` in *Bargmannia elongata*, `r length(Frillagalma_TF)` out of `r y[[4]] %>% sum` identified in *Frillagalma vityazi*, `r length(Agalma_TF)` out of `r y[[1]] %>% sum()` in *Agalma elegans* and `r length(Nanomia_TF)` out of `r y[[5]] %>% sum()` identified  in *Nanomia bijuga*). Many of these transcription factors are significantly differentially expressed in several zooids regardless of developmental stage (both mature and developing zooids), and a subset were found to be significantly differentially expressed only in particular zooids and developmental stages (\@ref(fig:transcription-factors)).

@@

summary %>% dplyr::mutate(
    Species= cell_spec(Species, italic= TRUE)
) %>% kable(caption="Number of genes per  species (library size), number of sampled zooids (at least two  replicates each, including developing and mature stages), total number of genes in gene trees (there can be multiple genes in a gene tree), and number of unique gene trees containing genes from this species.", caption.short ="Summary statistics for each of the sampled species, with details on the total number of genes in gene trees, and number of unique gene trees containing genes from this species.",booktabs=TRUE, align = c("l","c","c","c","c"), escape = FALSE) %>% kable_styling(full_width = F, font_size = 11) %>%
  kableExtra::kable_styling(latex_options = "hold_position") %>%
  column_spec(2:5, width = "5em")

@@

return(p)
}
```
In order to compare expression patterns for each zooid across siphonophore species, we used transcriptome and genome data from `r species_tree$tip.label %>% length()` cnidarian species to generate a total of `r gene_trees %>% length()` gene trees, of which `r gene_trees_ace_calibrated[!is.na(gene_trees_ace_calibrated)] %>% length()` gene trees passed filtering criteria. We used Orthofinder [@emms2019orthofinder] to reconcile the species tree and the gene trees and annotate each gene tree node as either a speciation or duplication events (Fig. \@ref(fig:methods-comp), step 1). Pruned and unpruned calibrated gene tree files are available as supplementary data. The number of genes represented in these gene trees is shown in table \@ref(tab:summarize-libraries). The internal nodes on these gene trees consist of `r nodes_ace %>% filter(is.na(species) & Ev=="S") %>% nrow()` speciation events and `r nodes_ace %>% filter(is.na(species) & Ev=="D") %>% nrow()` duplication events. Expression values, as transcripts per million 10K (TPM10K, see methods), were mapped to the tips of the gene trees (Fig. \@ref(fig:methods-comp), step 2), and maximum likelihood trait values were reconstructed at the nodes (Fig. \@ref(fig:methods-comp), step 3). Changes in expression across branches were calculated based on the difference between parent and child nodes, scaled by the branch length (Fig. \@ref(fig:methods-comp), step 4) [@dunn2018pairwise]. We focused on expression in a subset of zooids and tissues that are common across the sampled species: gastrozooids (developing and mature), nectophores (developing), palpons (mature), male and female gonodendra, and the pneumatophore. An exemplar gene tree, for Wnt family genes, with expression values from two zooids mapped at the nodes and tips is shown in figure \@ref(fig:wnt-plot).

@@

For comparitive expression analyses, expression values were normalized using a new method we call transcripts per million 10K (TPM10K). For gene $i$ of a given species, TPM is typically calculated as [@li2009rna]:

@@

This work was supported by the National Science Foundation (DEB-1256695 and the Waterman Award). CM was also supported in part by a RI-EPSCoR Fellowship, NSF EPS-1004057. Analyses were conducted with computational resources and services at the Center for Computation and Visualization at Brown University, supported in part by the NSF EPSCoR EPS-1004057 and the State of Rhode Island. Analyses were also conducted at the Yale Center for Research Computing (YCRC).
